Inefficient Concessions and Mediation

### 2021-02-21 draft.tex
add Ben's revisions to intro/literature

In the course of a military conflict, settlement negotiations often occur simultaneously with continued fighting. If settlement can be achieved, lives are saved and destruction averted. Third-party mediators often participate in such negotiations, but the literature remains divided regarding how and under what conditions mediation can be effective (CHAIN CITE).  This paper sheds new light on the productive role mediators can play by examining their effect on the efficiency of the concessions that combatants can offer. In our context, the efficiency of a concession is determined by how much of the material value of the concession is intact when the recipient receives it. If the party who gives the concession destroys any of its material value, the concession is deemed to be less than fully efficient.

@@

Two archetypal situations demonstrate the significance and difficulty of efficient concession-making in war settlement.  The first is the case of disarmament and the second is the concession of control over strategically valuable territory.

@@

Control of strategically valuable territory can pose similar challenges in both civil and interstate wars. Even if transferring control of the territory creates the best chance for peace, its transfer often involves a substantial risk that the territory will provide military advantage in a future conflict.

@@

Walter (1997) observes empirically that, due to the difficulty of making efficient disarmament concessions, civil wars tend to end on the battlefield, rather than through negotiated settlement. Interstate wars over territory are similarly prone to violent, rather than negotiated, resolution (CITE). Can mediation offer a path to settlement in these challenging contexts?

@@

This demonstration of the incentives for inefficient concessions leads to a new explanation for the role of mediators in conflict resolution. Third party involvement is ubiquitous in conflict and mediation has been extensively studied in the literature, but still the question of third party effectiveness is intensely debated.  Some authors have concluded that mediation has little impact, (Bercovitch 1996;  Bercovitch and Langley 1993; Fortna 2003) while others insist that a correct analysis finds mediation playing a positive role in resolving conflict (Dixon 1996; Beardsley et al. 2006).  But the term mediation encompasses a broad range of actions and interactions, making any one-size-fits-all conclusion on the efficacy of mediation difficult. The type of mediation affects the negotiation outcome (Bercovitch 2000; Beardsley et al. 2006) and timing is important (Greig 2005a).

It is essential to focus more precisely on the specific actions and environments that make use of particular types of mediation, which is how we approach the question. This paper models an environment where the primary issue is trust, not uncertainty about an opponent's capabilities, and where a mediator is employed who has manipulative abilities. Mediation rarely occurs in the absence of violent conflict (Bercovitch 1996; Beardsley 2006). In this paper, parties are in conflict and can benefit from seeking mediation in order to overcome mistrust. Parties are more likely to seek mediation when they are entrenched in costly conflict (Bercovitch and Jackson 2001; Greig 2005b; Greig and Diehl 2006; Terris and Maoz 2005; Svensson 2008). Alternatively, parties may seek a mediator to justify difficult concessions and avoid angering domestic constituencies (Allee and Huth 2006, Beardsley 2010).

@@

The intuition for this addition to the model is that the country that receives the concessions can use them to build up their military or to build up their civil society. Using concessions to build military strength hurts the giver of the concessions in war; using the concessions to strengthen civil society helps the giver in peace. The Golan heights is such an example: if Syria is an ally of Israel, the Golan Heights is a reasonable concession to make, but if Syria later turns out to betray a trust, surrender of the strategic land would be disastrous for Israel. %Another example is ADD more examples

@@

@@

@@

>>>>

In the course of a military conflict, settlement negotiations often occur simultaneously with continued fighting. If settlement can be achieved, lives are saved and destruction averted. Third-party mediators often participate in such negotiations, but the literature remains divided regarding how and under what conditions mediation can be effective (e.g., Bercovitch 2000; Greig 2005a; Beardsley et al. 2006; Duursma 2020).  This paper sheds new light on the productive role mediators can play by examining their effect on the efficiency of the concessions that combatants can offer. In our context, the efficiency of a concession is determined by how much of the material value of the concession is intact when the recipient receives it. If the party who gives the concession destroys any of its material value, the concession is deemed to be less than fully efficient.

@@

Two archetypal situations illustrate the significance and difficulty of efficient concession-making in war settlement.  The first is the case of disarmament and the second is the concession of control over strategically valuable territory.

@@

Control of strategically valuable territory can pose similar challenges in both civil and interstate wars. Even if transferring control of strategically valuable territory creates the best chance for peace, its transfer often involves a substantial risk that the territory will provide military advantage in a future conflict.

@@

Walter (1997) observes empirically that, due to the difficulty of making efficient disarmament concessions, civil wars tend to end on the battlefield, rather than through negotiated settlement. Interstate wars over territory are similarly prone to violent, rather than negotiated, resolution (Hensel 1996). Can mediation offer a path to settlement in these challenging contexts?

@@

This demonstration of the incentives for inefficient concessions leads to a new explanation for the role of mediators in conflict resolution. Third party involvement is ubiquitous in conflict and mediation has been extensively studied in the literature, but still the question of third party effectiveness is intensely debated.  Some authors have concluded that mediation has little impact, (Bercovitch 1996;  Bercovitch and Langley 1993; Fortna 2003) while others insist that a correct analysis finds mediation playing a positive role in resolving conflict (Dixon 1996; Beardsley et al. 2006).  But the term mediation encompasses a broad range of actions and interactions, making any one-size-fits-all conclusion on the efficacy of mediation difficult. The type of mediation affects the negotiation outcome (Bercovitch 2000; Beardsley et al. 2006), as does the legitimacy of the mediator (e.g., Duursma 2020) and the timing of the mediator's intervention (Greig 2005a).

It is essential to focus more precisely on the specific actions and environments that make use of particular types of mediation, which is how we approach the question. This paper models an environment where the primary issue is trust, not uncertainty about an opponent's capabilities, and where a mediator is employed who has manipulative abilities. Mediation rarely occurs in the absence of violent conflict (Bercovitch 1996; Beardsley 2006). In this paper, parties are in conflict and can benefit from seeking mediation in order to overcome mistrust. Parties are more likely to seek mediation when they are entrenched in costly conflict (Bercovitch and Jackson 2001; Greig 2005b; Greig and Diehl 2006; Terris and Maoz 2005; Svensson 2008). Alternatively, parties may seek a mediator to justify difficult concessions and avoid angering domestic constituencies (Allee and Huth 2006, Beardsley 2010; Beardsley and Lo 2014).

@@

The intuition for this addition to the model is that the country that receives the concessions can use them to build up their military or to build up their civil society. Using concessions to build military strength hurts the giver of the concessions in war; using the concessions to strengthen civil society helps the giver in peace. The Golan heights is such an example: if Syria maintains peaceful relations with Israel, the Golan Heights is a reasonable concession to make, but if Syria renews military conflict with Israel, surrender of the strategic land would be disastrous for Israel. %Another example is ADD more examples

@@

\item Beardsley, Kyle and Nigel Lo. (2014). ``Third-Party Conflict Management and the Willingness to Make Concessions." \emph{The Journal of Conflict Resolution} 58:2, 363-392.

@@

\item Duursma, Allard. (2020). ``African Solutions to African Challenges: The Role of Legitimacy in Mediating Civil Wars in Africa." \emph{International Organization} 74:2, 295-330.

@@

\item Hensel, Paul. (1996). ``Charting A Course To Conflict: Territorial Issues and Interstate Conflict, 1816-1992," \emph{Journal of Conflict Resolution} 15:1, 43-73.
